---
title: "Visualization 03 Challenge Complete"
author: "Maya Gann-Bociek & Matthew Schuelke, Ph.D."
date: '(`r format(Sys.time(), "%B %d, %Y")`)'
output:
  github_document: default
  html_notebook: default
---

## Introduction

We will continue to investigate NHANES data with "nhanes_c3.csv". 

## Tasks

1. Load the dependencies
2. Load the dataset 'nhanes_c3.csv' into your environment
3. Explore the data
4. Plot the data

## Load Dependencies

The following code loads the package dependencies for our analysis:

```{r package-load-1}
# tidyverse packages
library(dplyr)   
library(tidyr)   
library(ggplot2)  
library(readr)    
library(here)
library(skimr)
```

We now have the packages needed for our analyses.

## Load Data 

Load in data "nhanes_c3.csv"

```{r}
nhanes <- read_csv(here('data', 'nhanes_c3.csv'))
```

## Explore the Data

Print and summary the dataset to get familar with your data

```{r}
print(nhanes)
skim(nhanes)
```

## Plotting

Plot blood pressure ('BP') against 'ageMonths', in the following plotting aesthetics:

* Map type to color 
* Map gender to shape 

```{r}
ggplot(nhanes, aes(x = ageMonths, 
                   y = BP, 
                   color = type, 
                   shape = gender)) +
  geom_point()
```

Create a bar plot

* display the number of person in each 'AgeDecade'
* display the proportation of female and male with different colors

```{r}
ggplot(nhanes, aes(x = ageDecade, 
                   fill = gender)) + 
  geom_bar(stat = "count") +
  labs(x = "Age Decade",
       y = "Count") + 
  scale_fill_discrete("Gender", 
                      labels = c("Female", "Male"))
  
```
